cardinalR: Generating Interesting High-Dimensional Data Structures

### 2025-12-15 paper-cardinalR.Rmd
clustering example refined

<!--script/05_gen_cluster_stat.R-->
```{r, label = "fig-cluster-stats", fig.width=10, fig.height=3, out.width="100%", layout="l-body", fig.cap="Cluster validity metrics for solutions with $2–10$ clusters obtained using $k$-means, hierarchical, and model-based clustering. Several indices consistently suggest that $4–5$ clusters provide the best balance of separation and compactness, with $k$-means performing slightly better across metrics."}

@@

Figure \@ref(fig:fig-cluster-stats) shows a selection of cluster metrics for $2-10$ clusters for each of the methods, $k$-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a $4-5$ cluster solution. The Calinski–Harabasz index increases sharply up to four or five clusters before levelling off, the Dunn and Pearson gamma indices attain local maxima in this range, and both the silhouette width and within-cluster sum of squares show diminishing improvement beyond five clusters. Together, these patterns indicate that $4-5$ clusters provide a good balance of separation and compactness, with $k$-means performing slightly better than the other methods across most metrics.

```{r}
## k-means with  four clusters
kcl4 <- kmeans(five_clusts[, -5], centers = 4, nstart = 20)$cluster

## k-means with  five clusters
kcl5 <- kmeans(five_clusts[, -5], centers = 5, nstart = 20)$cluster
```

@@

`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:highd-data-clusters-algo-html)"} else { "Figure \\@ref(fig:highd-proj-clust-algo-pdf)"})` shows the four- and five-cluster $k$-means solutions further clarify why these partitions do not fully identify the underlying structure. The predicted clusters show noticeable overlap, with colours repeatedly blending as the tour moves; evidence that points from different true clusters are assigned to the same cluster. The $4$-cluster solution shows large, diffuse regions where several true clusters are merged, while the $5$-cluster solution partially separates some structures but still misclassifies substantial portions of multiple clusters. These patterns reflect the mismatch between the helical spherical shape present in the data and the spherical decision boundaries imposed by $k$-means. Even though the cluster statistics favour a $4-5$ cluster solution overall, `r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:highd-data-clusters-algo-html)"} else { "Figure \\@ref(fig:highd-proj-clust-algo-pdf)"})` reveals that neither solution provides clear separation, highlighting the importance of visual diagnostics when interpreting clustering results.

>>>>

<!--script/05_gen_cluster_stat.R-->
```{r, label = "fig-cluster-stats", fig.width=8, fig.height=6, out.width="100%", layout="l-body", fig.cap="Cluster validity metrics for solutions with $2–10$ clusters obtained using $k$-means, hierarchical, and model-based clustering. Several indices consistently suggest that $4–5$ clusters provide the best balance of separation and compactness, with $k$-means performing slightly better across metrics."}

@@

Figure \@ref(fig:fig-cluster-stats) shows a selection of cluster metrics for $2-10$ clusters for each of the methods, $k$-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a $4-5$ cluster solution. Pearson gamma (`pearsongamma`) increases sharply up to five clusters before levelling off, Calinski–Harabasz index (`ch`) increases sharply from 4 to 5 clusters and Dunn (`dunn2`) has a maximum at 5 for two methods and at 4 for $k$-means. All of these are interpreted as higher is better. With the other three, lower is better. WB ratio (`wbratio`) and within-cluster sum of squares (`within.cluster.ss`) steadily decline with number of clusters, possibly elbowing around 5 clusters. The S-index (`sindex`) is optimised at 4 clusters for $k$-means, 3, 6 or 8 for hierarchical clustering, and 4 or 8 for model-based. Overall, $k$-means performs slightly better than the hierarchical and model-based clustering across most metrics and number of clusters.

```{r}
set.seed(1214)

## k-means with  four clusters
kcl4 <- kmeans(five_clusts[, -5], centers = 4, nstart = 20)$cluster

## k-means with  five clusters
kcl5 <- kmeans(five_clusts[, -5], centers = 5, nstart = 20)$cluster

@@

`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:highd-data-clusters-algo-html)"} else { "Figure \\@ref(fig:highd-proj-clust-algo-pdf)"})` shows the four- and five-cluster $k$-means solutions, with cluster id used to colour the points. Neither solution captures the geometric nature of the true clusters, but they are both reasonable partitions of the data. To examine either one, it is best to subset to a single cluster to view in the tour. With each solution, the five original shapes are each split by the clustering. More than 5 clusters would be needed to better capture the original shapes.

### 2025-12-15 paper-cardinalR.tex
clustering example refined

Figure \ref{fig:fig-cluster-stats} shows a selection of cluster metrics for \(2-10\) clusters for each of the methods, \(k\)-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a \(4-5\) cluster solution. The Calinski--Harabasz index increases sharply up to four or five clusters before levelling off, the Dunn and Pearson gamma indices attain local maxima in this range, and both the silhouette width and within-cluster sum of squares show diminishing improvement beyond five clusters. Together, these patterns indicate that \(4-5\) clusters provide a good balance of separation and compactness, with \(k\)-means performing slightly better than the other methods across most metrics.

@@

Figure \ref{fig:highd-proj-clust-algo-pdf} shows the four- and five-cluster \(k\)-means solutions further clarify why these partitions do not fully identify the underlying structure. The predicted clusters show noticeable overlap, with colours repeatedly blending as the tour moves; evidence that points from different true clusters are assigned to the same cluster. The \(4\)-cluster solution shows large, diffuse regions where several true clusters are merged, while the \(5\)-cluster solution partially separates some structures but still misclassifies substantial portions of multiple clusters. These patterns reflect the mismatch between the helical spherical shape present in the data and the spherical decision boundaries imposed by \(k\)-means. Even though the cluster statistics favour a \(4-5\) cluster solution overall, Figure \ref{fig:highd-proj-clust-algo-pdf} reveals that neither solution provides clear separation, highlighting the importance of visual diagnostics when interpreting clustering results.

>>>>

Figure \ref{fig:fig-cluster-stats} shows a selection of cluster metrics for \(2-10\) clusters for each of the methods, \(k\)-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a \(4-5\) cluster solution. Pearson gamma (\texttt{pearsongamma}) increases sharply up to five clusters before levelling off, Calinski--Harabasz index (\texttt{ch}) increases sharply from 4 to 5 clusters and Dunn (\texttt{dunn2}) has a maximum at 5 for two methods and at 4 for \(k\)-means. All of these are interpreted as higher is better. With the other three, lower is better. WB ratio (\texttt{wbratio}) and within-cluster sum of squares (\texttt{within.cluster.ss}) steadily decline with number of clusters, possibly elbowing around 5 clusters. The S-index (\texttt{sindex}) is optimised at 4 clusters for \(k\)-means, 3, 6 or 8 for hierarchical clustering, and 4 or 8 for model-based. Overall, \(k\)-means performs slightly better than the hierarchical and model-based clustering across most metrics and number of clusters.

@@

Figure \ref{fig:highd-proj-clust-algo-pdf} shows the four- and five-cluster \(k\)-means solutions, with cluster id used to colour the points. Neither solution captures the geometric nature of the true clusters, but they are both reasonable partitions of the data. To examine either one, it is best to subset to a single cluster to view in the tour. With each solution, the five original shapes are each split by the clustering. More than 5 clusters would be needed to better capture the original shapes.
